---
title: Writing challenge Day 14
author: "Matias Andina"
date: 2018-09-13
tags: ["research", "writing", "learning", "DocumentingLearning"]

---

# Second Week

This week I had a lot of late working days. It has been certainly tiresome. As a consequence, I was less able to write, got home really tired. 

I still have that voice powering my desire to write. Actually, over the weekend I finished a short story that I had in my list for at least a year. That achievement feels really good and pumps the desire to keep writing and, most importantly, *finishing* more projects than those I start.

## The data

The word count has been erratic. This week, I have had a couple of days in which I wasn't even able to sit down to write. However, I think it has a somewhat consistent trend around a global mean (~500 words per bout). I have high hopes for improvement of this `word-count-per-bout` metric. 

```{r, echo=FALSE, results="hold", message=F, warning=FALSE}
library(ggplot2)
library(dplyr)
suppressPackageStartupMessages(library(lubridate))
```


```{r echo=FALSE, message=FALSE, warning=FALSE}

tracking_writing <- read.csv("C:/Users/Matias/Dropbox/mati/Escritos/Cuentos/tracking_writing.csv")

tracking_writing <- dplyr::select(tracking_writing, -place) %>%
  mutate(date=ymd(date))

ggplot(tracking_writing, aes(date,word_count))+
  geom_point()+
  scale_x_date(date_labels="%b %d",date_breaks  ="1 day")+
  geom_smooth(color="black")+
  theme_bw()+
  ggtitle("Word count per writing bout")
  

```

If we look at the total writing per day, the data looks a little bit different. 

```{r, echo=FALSE, warning=FALSE, message=FALSE}

tracking_writing %>% group_by(date) %>%
  summarise(total_count = sum(word_count)) %>%
ggplot(aes(date,total_count))+
  geom_point()+
  scale_x_date(date_labels="%b %d",date_breaks  ="1 day")+
  geom_smooth(color="black")+
  theme_bw()+
  ggtitle("Total count per day")+
  geom_hline(yintercept = 2000, lty=2)+
  annotate(geom="text", label="daily target", x= ymd("2018-09-11"), y=2100)


```

I think that 2000 words a day is a pretty high standard, not only because it's a high number of words, but holding a consistent mark of 2000 words requires you to spend a significant amount of time writing every day. Although I think writing this much is doable, I don't think it's possible to maintain the high level if it's not part of your job (hence you devote a specific amount of time each day to do it). I have not included the words I've written in *job related* writing in this count, only leisure writing. Incorporating the mindset behind this challenge to a job that needs the production of text in massive amounts can definitely produce good results. Successful PIs are machines that are able to write grants and papers for a living (and the improvement of science?) in a consistent manner, a mix of playing the numbers and exceptional management of scientific endeavors.    

## The distractions

Distractions while writing are a well-known issue. I have not been able to turn off distractions (e.g., internet, messages). I am mostly free of social media distractions. However, I have not committed to turn off my real life events in order to write. I attempt (and yes, fail pretty often) to schedule around my life in order to sit, focus and write.

## The good

Finishing a short story feels really good. I have also been having ideas that can work in other stories and improve the plot forward. Whenever I encountered a block, I made notes to myself and moved on, trying to advance with another project while my mind works on the background, in order to solve the problem that's preventing me from writing a particular section.

Some days I have 2 bouts of writing. It's either because I move from the living table to the bed or because I wrote a bit during the morning and then came back to it during the afternoon/night. Whenever I do sit to write I have a sense of improved flow. I am able just let ideas transform into words with ease. The production rate is not quite up yet, mostly due to bad planning, lack of ideas or lack of clear outlines that help generating them. 

## To improve

I still need to work on the outline. I am starting to notice the stitches in the books that I read. It is pretty clear when authors use outlines and develop ideas into short paragraphs. Although it is definitely a great way to put many ideas into text and increase the word count, I believe it can be a dangerous way to write, you risk repeating yourself too much (e.g., developing ideas in a new chunk/paragraph that were developed 17 chunks before).   

*Note to self: You are not suposed to develop the previous idea that far while doing the outline, continue outlining please.*

***