---
title: "Github style calendar heatmaps"
author: "Matias Andina"
date: 2018-05-27
tags: ["research", "design", "R"]
---

I like how the commit heatmap looks in Github. I wanted to play with something that could be plotted that way. I've seen some beautiful things done in `d3` and javascript. But, of course, I wanted to make it in `R`. Turns out a bunch of other people have great ideas for how to go about it. Thus, I'm borrowing heavily from them[^1].


```{r setup, message=F, warning=F}

library(ggplot2)
library(dplyr)
library(lubridate)
library(viridis)  # Color palette
library(ggthemes) # theme tufte


```

Let's generate a `data.frame` for May 2018.


```{r}

# choose dates
start_date <- ymd("2018-05-01")
end_date <- ymd("2018-05-31")

d <- data_frame(
    date = seq(start_date, end_date, by = "days"),
    month = month(date),
    year = format(date, "%Y"),
    week = as.integer(format(date, "%W")) + 1,  # Week starts at 1
    day = factor(weekdays(date, T), 
                 levels = rev(c("Mon", "Tue", "Wed", "Thu",
                                "Fri", "Sat", "Sun"))))


```


This is how our data looks like now

```{r data-preview}

head(d)

```

Now, let's assume I registered some events in my life and that my data looks something like: A date column (*date*), and the number of events that happened on a particular date (*n*).

```{r toy-data, echo=F}

df <- structure(list(date = structure(c(17666, 17667, 17668, 17669, 
17670, 17671, 17672, 17673, 17674, 17675, 17676, 17677, 17678
), class = "Date"), n = c(1L, 1L, 1L, 2L, 4L, 2L, 2L, 2L, 2L, 
5L, 2L, 1L, 4L)), .Names = c("date", "n"), class = c("tbl_df", 
"tbl", "data.frame"), row.names = c(NA, -13L))

```


```{r mydata}

df

```


I can join both `data.frames` and visualize!  

```{r line-plot, warning=FALSE}

df_plot <- d %>% left_join(df, by = "date") 

df_plot %>%
  mutate(n=ifelse(is.na(n), 0, n)) %>% ## Fill the NAs with zeros
  ggplot(aes(date, n)) +
  geom_line(lwd=0.7)+
  geom_point(size=2, shape=21, fill="black", colour="white", stroke=2)+
    theme_classic() +
  theme(panel.background = element_rect(colour = "black"))+
  ylab("Number of events")

```

My goal is not to analyze long term trends like seasonality. Thus, this plot is rather unremarkable. Not only because it is a small toy-like dataset, but because it fails to inform calendar information. Let's try to make it better!

```{r}

calendar_plot <- function(data, color.scale = "viridis",
                          viridis.pal = "D", dir = 1){
    
    p <- ggplot(data, aes(x = week, y = day, fill = n)) +
                geom_tile(color = "white", size = 0.8) +
        facet_wrap("year", ncol = 1) +
        theme_tufte() +
        theme(axis.ticks = element_blank(),
              legend.position = "bottom",
              legend.key.width = unit(1, "cm"),
              strip.text = element_text(hjust = 0.01,
                                        face = "bold", size = 12),
              text = element_text(size=16)) + 
      ylab("")
    
    
    
    # Let's add more than one possible pallete. Default keeps being viridis
    # Add case switch? or add 'none' for user to define their own ?
    
   if(color.scale=="viridis"){
  
     
    p <- p + scale_fill_viridis(name="Number of Events", 
                           option = viridis.pal,  # Variable color palette
                           direction = dir,  # Variable color direction
                           na.value = "grey93",
                           limits = c(1, max(data$n)))
  
   } else if(color.scale == 'greens') {
     
      p <- p + scale_fill_gradient(name="Number of Events",
                                   low="lightyellow2", high="darkgreen", 
                                   na.value = "grey93")
   } else{
     
     error("Accepted color.scale are 'viridis' and 'greens'")
     
   }

    
    ## x axis control of labels
    
    
    num_months <- length(unique(data$month))
    
    if(num_months > 1){
      
      p <- p + scale_x_continuous(
            expand = c(0, 0),
            breaks = seq(1, 52, length = 12),
            labels = c("Jan", "Feb", "Mar", "Apr", "May", "Jun",
                       "Jul", "Aug", "Sep", "Oct", "Nov", "Dec"))

    } else {
      
      # do nothing
      
      p <- p + xlab("Week Number.")
      
    }
        
      
    
    print(p)
}


```



We can use `calendar_plot()` function now to make a plot in *calendar-like* shape. It is easier to see the data, even with such as small dataset. Below there are two color scale representations of the same data.

```{r calendar-plot}

calendar_plot(df_plot, 'greens')

calendar_plot(df_plot, viridis.pal = "B")

```


***

## UPDATE

I was curious about how data would look like for a longer span. Here's the data for a longer time interval.  

```{r updated-data, echo=FALSE}

df <- read.csv(file = 'data_for_posts/2018-05-27_longer_data.csv', stringsAsFactors = FALSE)

df$date <- ymd(df$date)

# choose dates
start_date <- ymd("2018-05-01")
end_date <- ymd("2018-08-31")

d <- data_frame(
    date = seq(start_date, end_date, by = "days"),
    month = month(date),
    year = format(date, "%Y"),
    week = as.integer(format(date, "%W")) + 1,  # Week starts at 1
    day = factor(weekdays(date, T), 
                 levels = rev(c("Mon", "Tue", "Wed", "Thu",
                                "Fri", "Sat", "Sun"))))

df_plot <- d %>% left_join(df, by = "date") 


```


```{r update-plot}

calendar_plot(df_plot)

```

When I did this update, I realized that my `calendar_plot()` function should handle internally the creation of the `data.frame` named `d` that serves as a placeholder. I guess that will happen in a following update `:)`.  

***
[^1]: Great resources  [here](https://vuorre.netlify.com/post/2016/2016-03-24-github-waffle-plot/), [and here](http://ethen8181.github.io/Business-Analytics/articles/calendar_heatmaps/calendar_heatmaps.html)

